%% LaRT Clump Medium Technical Description
%% Author: Kwang-Il Seon
\documentclass[english,a4paper]{kiseon_memo}
\synctex=1
\usepackage{listings}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{microtype}

\lstset{
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue}\bfseries,
  commentstyle=\color{gray},
  stringstyle=\color{olive},
  frame=single,
  breaklines=true,
  numbers=left,
  numberstyle=\tiny\color{gray},
  language=[95]Fortran
}

\usepackage{babel}
\newcommand{\fdg}{\hbox{$.\!\!^\circ$}}
\begin{document}

\title{Clumpy Medium Extension of LaRT: Algorithm Description}
\author{Kwang-Il Seon}
\date{2026-04-26}
\maketitle

\tableofcontents
\bigskip

%=============================================================================
\section{Overview}
%=============================================================================

The clumpy medium mode adds a population of $N$ non-overlapping spherical
clumps inside a spherical domain of radius $R$.  Each clump is a sphere of
radius $r_c$ with uniform Lyman-$\alpha$ opacity; the inter-clump medium is
vacuum.

The mode is activated by
\begin{lstlisting}
par%use_clump_medium = .true.
\end{lstlisting}
\begin{sloppypar}
in the input namelist.  All raytrace procedure pointers are redirected to
clump-aware routines (\texttt{raytrace\_clump.f90}) while scatter and
peel-off reuse the standard Cartesian routines from
\texttt{scattering\_car.f90} and \texttt{peelingoff\_rect.f90}, since the
uniform clump physical parameters ($T$, $v_\mathrm{bulk}$) are broadcast to
the Cartesian grid arrays (\texttt{grid\%Dfreq}, \texttt{grid\%voigt\_a}).
\end{sloppypar}

New source files introduced for this mode:
\begin{center}
\begin{tabular}{ll}
\hline
File & Role \\
\hline
\texttt{clump\_mod.f90}       & Clump population: init, RSA placement, CSR grid, raytrace helpers \\
\texttt{raytrace\_clump.f90}  & Raytrace through the clump medium (DDA + sphere intersection) \\
\texttt{grid\_mod\_clump.f90} & Grid setup: creates Cartesian shell grid, then calls \texttt{init\_clumps} \\
\hline
\end{tabular}
\end{center}

%=============================================================================
\section{Physical Setup}
%=============================================================================

\subsection{Geometry}

\begin{sloppypar}
The outer domain is a sphere of radius $R = \texttt{par\%rmax}$ (code units).
$N$ spherical clumps of radius $r_c = \texttt{par\%clump\_radius}$ are placed
at random inside this sphere (non-overlapping).
\end{sloppypar}

A Cartesian grid of $N_x \times N_y \times N_z$ cells is created with the
same bounding box $[-R, R]^3$.  Grid arrays ($\texttt{rhokap}$, $\texttt{Dfreq}$,
$\texttt{voigt\_a}$, $\texttt{vf}$) are set to uniform values:
\begin{itemize}
  \item $\texttt{rhokap} = 0$ everywhere (opacity lives only inside clumps);
  \item $\texttt{Dfreq} = \Delta\nu_D^{cl}$ (clump Doppler frequency);
  \item $\texttt{voigt\_a} = a_V^{cl}$ (clump Voigt parameter);
  \item $\texttt{vfx} = \texttt{vfy} = \texttt{vfz} = 0$.
\end{itemize}
Peel-off and scattering routines read these uniform grid values, so no
special-casing is needed inside those modules.

\subsection{Clump Count}

The user specifies one of three equivalent parameters; the code derives $N$
from whichever is set:

\paragraph{From $f_\mathrm{cov}$ (covering factor):}
The mean number of clumps intercepted by a radial ray from the centre to the
sphere surface is
\begin{equation}
  f_\mathrm{cov} = N\,\frac{\pi r_c^2}{\pi R^2}\cdot\frac{R}{\frac{4}{3}\pi R^3/(4\pi R^2)}
  = \frac{3}{4}\,N\!\left(\frac{r_c}{R}\right)^2 ,
\end{equation}
hence
\begin{equation}
  N = \frac{4}{3}\,f_\mathrm{cov}\!\left(\frac{R}{r_c}\right)^2 .
\end{equation}
Input parameter: \texttt{par\%clump\_f\_cov}.

\paragraph{From $f_\mathrm{vol}$ (volume filling factor):}
\begin{equation}
  f_\mathrm{vol} = N\left(\frac{r_c}{R}\right)^3 ,
  \qquad N = f_\mathrm{vol}\!\left(\frac{R}{r_c}\right)^3 .
\end{equation}
Input parameter: \texttt{par\%clump\_f\_vol}.

\paragraph{Direct specification:} \texttt{par\%clump\_N\_clumps}.

\subsection{Clump Physical Parameters}

The clump temperature $T_c$ defaults to $\texttt{par\%temperature}$ unless
\texttt{par\%clump\_temperature} $> 0$.  Derived quantities:
\begin{align}
  v_\mathrm{th} &= v_{\mathrm{th},1}\sqrt{T_c},\quad
                   v_{\mathrm{th},1} = \sqrt{2k_B/m_H}\approx 12.84\;\mathrm{km\,s^{-1}/\sqrt{10^4\,K}},\\
  \Delta\nu_D^{cl} &= v_\mathrm{th} / (\lambda_0 \times \mathrm{km2um}),\\
  a_V^{cl} &= \frac{A_{21}}{4\pi\,\Delta\nu_D^{cl}},
\end{align}
where $\lambda_0$ is the Lyman-$\alpha$ rest wavelength.

Three mutually exclusive parameters specify the opacity of each individual
clump (the first positive value in the priority order wins):

\paragraph{1.~\texttt{par\%clump\_tau0} — line-centre optical depth (dimensionless):}
The line-centre $\tau$ of one clump from its centre to its surface.
\begin{equation}
  \kappa_c = \frac{\tau_0}{\phi(0,a_V^{cl})\, r_c},
\end{equation}
where $\phi(x, a)$ is the Voigt profile at dimensionless frequency $x$.

\paragraph{2.~\texttt{par\%clump\_NHI} — per-clump HI column density [cm$^{-2}$]:}
The HI column density of a \emph{single clump} from its centre to its surface.
This must be clearly distinguished from the \emph{total system} column density
(\texttt{par\%N\_HImax}), which is the average HI column density along a
radial sightline through the entire clumpy sphere (centre to outer boundary,
averaged over all sightlines through the random clump population).

Given $N_{\rm HI}^{cl} = n_H\,r_c\,d_{\rm cm}$, the opacity follows as
\begin{equation}
  \kappa_c = \frac{N_{\rm HI}^{cl}\,\sigma_0}{r_c},
\end{equation}
where $\sigma_0 \equiv \texttt{line\%cross0}/\Delta\nu_D^{cl}$ [cm$^2$] is the
line-centre cross-section.
(\texttt{line\%cross0} $= \pi e^2 f/(m_e c\sqrt{\pi})$ has units [cm$^2$\,Hz];
dividing by $\Delta\nu_D^{cl}$ [Hz] gives the physical cross-section [cm$^2$].)
This is equivalent to $\tau_0 = N_{\rm HI}^{cl}\,\sigma_0\,\phi(0,a_V^{cl})$.
Unlike \texttt{clump\_nH}, this option does \emph{not} require
\texttt{par\%distance\_unit}.

\paragraph{3.~\texttt{par\%clump\_nH} — hydrogen number density [cm$^{-3}$]:}
$\kappa_c = n_H\,\sigma_0\,d_\mathrm{cm}$,
where $\sigma_0 \equiv \texttt{line\%cross0}/\Delta\nu_D^{cl}$ [cm$^2$]
(requires \texttt{par\%distance\_unit}).

\subsection{Derived Global Optical-Depth Quantities ($\tau_\mathrm{homo}$,
$N_\mathrm{HI,homo}$, $\tau_\mathrm{max}$, $N_\mathrm{HI,max}$)}
\label{subsec:tauhomo}

LaRT carries four scalar parameters that describe the line-centre
optical depth and HI column density of the medium as a whole:
\begin{description}
\item[\texttt{par\%tauhomo}, \texttt{par\%N\_gashomo}] — the line-centre
  optical depth and HI column density of an equivalent
  \emph{homogeneous} medium that contains the same total HI mass spread
  uniformly over the sphere volume.
\item[\texttt{par\%taumax}, \texttt{par\%N\_gasmax}] — the line-centre
  optical depth and HI column density along a single radial sightline
  from the centre to the $+z$ surface.
\end{description}
For a homogeneous medium these four numbers are equal up to geometric
factors; for a clumpy medium \texttt{taumax} along any specific sightline
fluctuates strongly (it can be zero if no clumps lie along the LOS, or
many times the homogeneous value if several clumps are stacked), while
\texttt{tauhomo} is a deterministic property of the medium (mass / volume).

The auto-detection of the simulation frequency range
($\texttt{par\%xfreq\_max}/\texttt{xfreq\_min}$) inside
\texttt{car\_setup\_freq\_grid} uses both quantities:
\begin{equation}
  a_{\tau,3} = \left(a_V \cdot \tau_\mathrm{homo}\right)^{1/3},\qquad
  x_\mathrm{max} \approx x_\mathrm{scale}(\tau_\mathrm{max}) \cdot a_{\tau,3}.
\end{equation}
Hence even when the user supplies \texttt{velocity\_min}/\texttt{velocity\_max}
explicitly, having physically meaningful \texttt{tauhomo} and \texttt{taumax}
is important for the FITS header to be informative and for the
fall-back auto-range to produce a usable grid.

\paragraph{Derivation in clump mode.}
\texttt{init\_clumps()} is called BEFORE \texttt{grid\_create()} inside
\texttt{grid\_create\_clump}, so the realised clump count $N_c$, opacity
$\kappa_c$, and Voigt parameter $a_V^{cl}$ are all known when the global
$\tau$/N values are derived.  The volume filling factor and the LaRT
covering factor (defined as the expected number of clumps along a radial
path of length $R$) are
\begin{equation}
  f_\mathrm{vol} = N_c \left(\frac{r_c}{R}\right)^3,\qquad
  f_\mathrm{cov} = \frac{3}{4}\,N_c \left(\frac{r_c}{R}\right)^2.
\end{equation}
The line-centre optical depth of one clump (centre to surface) is
$\tau_0 = \kappa_c\,\phi(0,a_V^{cl})\,r_c$.  Distributing the same total
HI mass uniformly over the sphere volume gives the homogeneous-equivalent
opacity per unit length
$\langle\kappa\rangle_\mathrm{lc} = f_\mathrm{vol}\,\kappa_c\,\phi(0,a_V^{cl})$,
and the volume-averaged $\tau$ from centre to surface is therefore
\begin{equation}
  \tau_\mathrm{homo} = \langle\kappa\rangle_\mathrm{lc}\cdot R
                     = f_\mathrm{vol}\,\kappa_c\,\phi(0,a_V^{cl})\,R
                     = \tfrac{4}{3}\,f_\mathrm{cov}\,\tau_0.
\end{equation}
The same value is also the EXPECTED radial $\tau_\mathrm{max}$ for a
random sightline through the clumpy sphere (so the two have the same
expectation for clumpy media, and are identical for the truly homogeneous
limit).  The HI columns are obtained by replacing $\tau_0$ with the
per-clump column $N_\mathrm{HI}^{cl}$:
\begin{equation}
  N_\mathrm{HI,homo} = N_\mathrm{HI,max}^\mathrm{(expected)}
    = \tfrac{4}{3}\,f_\mathrm{cov}\,N_\mathrm{HI}^{cl},\qquad
  N_\mathrm{HI}^{cl} = \frac{\kappa_c}{\sigma_0} = \frac{\kappa_c\,\Delta\nu_D^{cl}}{\texttt{line\%cross0}}\;r_c.
\end{equation}
For the canonical \texttt{clump\_NHI = 7.5e17 cm$^{-2}$, $f_\mathrm{cov} = 1$, $T = 10^4$ K} test
case this gives $N_\mathrm{HI,homo} = 1.0\times10^{18}\,\mathrm{cm}^{-2}$,
$\tau_\mathrm{homo} = 5.9\times10^{4}$, and
$a_{\tau,3} = 3.0$.

The derivation only overrides parameters that the user did not specify
(stored as $\le 0$ before \texttt{read\_input}); explicit user values are
always preserved.

\subsection{Clump Velocities}

Each clump's velocity is the sum of two independent components:
\begin{equation}
  \mathbf{v}_i = \mathbf{v}_\mathrm{sys}(\mathbf{r}_i) + \mathbf{v}_\mathrm{rand,i},
\end{equation}
where $\mathbf{v}_\mathrm{sys}$ is a position-dependent systematic flow
(set by \texttt{par\%velocity\_type}) and $\mathbf{v}_\mathrm{rand}$ is an
independent Gaussian random component (set by \texttt{par\%clump\_sigma\_v}).
Either component can be zero.

\paragraph{Random component:}
If \texttt{par\%clump\_sigma\_v} $> 0$, each clump is assigned
$\mathbf{v}_{\mathrm{rand},i} \sim \mathcal{N}(0,\sigma_v^2)$ per component
(drawn in \texttt{generate\_clumps}).

\begin{sloppypar}
\paragraph{Systematic component (\texttt{par\%velocity\_type}):}
After RSA placement, \texttt{assign\_clump\_velocities\_from\_type} evaluates
the flow velocity at each clump centre and \emph{adds} it to the existing
(possibly non-zero) velocity.  Supported types:
\end{sloppypar}

\begin{center}
\begin{tabular}{lll}
\hline
\texttt{velocity\_type} & Formula & Parameters \\
\hline
\texttt{'hubble'} & $\mathbf{v} = V_\mathrm{exp}\,\mathbf{r}/R$ & \texttt{Vexp} \\
\texttt{'constant\_radial'} & $\mathbf{v} = V_\mathrm{exp}\,\hat{\mathbf{r}}$ & \texttt{Vexp} \\
\texttt{'parallel\_velocity'} & $\mathbf{v} = (V_x, V_y, V_z)$ & \texttt{Vx, Vy, Vz} \\
\texttt{'ssh'} & see below & \texttt{Vpeak, rpeak, DeltaV} \\
\texttt{'rotating\_solid\_body'} & $\mathbf{v} = V_\mathrm{rot}(-y,x,0)/R$ & \texttt{Vrot} \\
\texttt{'rotating\_galaxy\_halo'} & flat rotation about $z$ & \texttt{Vrot, rinner} \\
\hline
\end{tabular}
\end{center}

All velocity magnitudes are in km/s.  The \texttt{'ssh'} model (Song, Seon \&
Hwang 2020) gives a linear rise inside $r_\mathrm{peak}$ and a Hubble-like
outer flow:
\begin{equation}
  v(r) = \begin{cases}
    V_\mathrm{peak}\,r / r_\mathrm{peak} & r < r_\mathrm{peak}, \\
    V_\mathrm{peak} + \Delta V\,(r - r_\mathrm{peak})/(R - r_\mathrm{peak}) & r \ge r_\mathrm{peak}.
  \end{cases}
\end{equation}
The \texttt{'rotating\_galaxy\_halo'} model uses the cylindrical radius
$\rho = \sqrt{x^2+y^2}$ and gives solid-body rotation inside $r_\mathrm{inner}$
and a flat curve ($v_\phi = V_\mathrm{rot}$) outside.

%=============================================================================
\section{Clump Placement: Random Sequential Addition (RSA)}
%=============================================================================

\subsection{Concept}

Random Sequential Addition (RSA) is a simple stochastic algorithm for
generating configurations of non-overlapping hard spheres.  Candidate
positions are drawn at random one at a time; each is tested against all
previously accepted clumps and either accepted (no overlap) or discarded
irreversibly (overlap detected).  Because accepted positions are never moved,
RSA requires no iterative relaxation.

The non-overlap condition enforces a minimum centre-to-centre distance of
$2r_c$.  RSA reliably converges for volume filling factors $f_\mathrm{vol}
\lesssim 35\%$; above this threshold, the fraction of unblocked placement
space decreases so steeply that placing the last few clumps may require
enormous numbers of attempts.  The theoretical 3-D jamming limit for RSA is
$f_\mathrm{vol} \approx 38\%$.

\subsection{Uniform Sampling Inside the Sphere}

Candidate positions must be distributed \emph{uniformly} inside the outer
sphere of radius $R$.  This is achieved by box-rejection:
\begin{enumerate}
  \item Draw $x_c$, $y_c$, $z_c$ each uniformly from $[-R,\, R]$.
  \item Accept if $x_c^2 + y_c^2 + z_c^2 \le R^2$; otherwise draw again.
\end{enumerate}
The acceptance probability equals the sphere-to-cube volume ratio
$\pi/6 \approx 52\%$, so on average only $6/\pi \approx 1.91$ draws are
needed per accepted point.  Box-rejection avoids trigonometric functions and
yields an exactly uniform distribution in 3-D Cartesian space.

\subsection{Placement Algorithm}

\begin{lstlisting}
do while (icl < N_clumps)

  !--- Step 1: draw candidate inside sphere (box rejection)
  do
    xc = (2.0*rand() - 1.0) * R
    yc = (2.0*rand() - 1.0) * R
    zc = (2.0*rand() - 1.0) * R
    if (xc**2 + yc**2 + zc**2 <= R**2) exit
  enddo

  !--- Step 2: locate candidate in hash-grid cell (ig, jg, kg)
  ig = int((xc - xmin) / cell_size)
  jg = int((yc - ymin) / cell_size)
  kg = int((zc - zmin) / cell_size)

  !--- Step 3: check 27 neighbouring cells for overlap
  overlap = .false.
  outer: do kg2 = kg-1, kg+1
         do jg2 = jg-1, jg+1
         do ig2 = ig-1, ig+1
    if (out of bounds) cycle
    jnb = head(ig2, jg2, kg2)       ! first clump in this hash cell
    do while (jnb > 0)
      if (dist2(xc,yc,zc, cl_x(jnb),cl_y(jnb),cl_z(jnb)) < (2*rc)**2) then
        overlap = .true.; exit outer ! reject immediately on first overlap
      endif
      jnb = nxt(jnb)                ! follow linked list to next clump
    enddo
  enddo; enddo; enddo outer

  if (overlap) cycle                ! reject: draw new candidate

  !--- Step 4: accept -- record position and random velocity
  icl = icl + 1
  cl_x(icl) = xc;  cl_y(icl) = yc;  cl_z(icl) = zc
  if (sigma_v > 0.0) then
    cl_vx(icl) = sigma_v * gauss_random()  ! one independent draw per component
    cl_vy(icl) = sigma_v * gauss_random()
    cl_vz(icl) = sigma_v * gauss_random()
  endif

  !--- Step 5: prepend clump to the linked list for cell (ig, jg, kg)
  nxt(icl)         = head(ig, jg, kg)
  head(ig, jg, kg) = icl

enddo
\end{lstlisting}

\subsection{Linked-List Hash Grid and the 27-Cell Rule}
\label{sec:rsa_hash}

Naively, checking candidate $n$ against all $n-1$ accepted clumps costs
$O(n)$, giving $O(N^2)$ total work.  The linked-list hash grid reduces each
overlap check to $O(1)$ expected cost by spatially indexing accepted clumps.

The domain $[-R, R]^3$ is partitioned into $N_g^3$ uniform cells with
\begin{equation}
  N_g = \min\!\bigl(512,\; \lfloor N^{1/3} \rfloor + 1\bigr),
  \qquad
  \ell_\mathrm{RSA} = \frac{2R}{N_g}.
  \label{eq:ng}
\end{equation}
Two integer arrays encode the linked lists: \texttt{head(i,j,k)} is the index
of the first clump in cell $(i,j,k)$ (0 if empty), and \texttt{nxt(icl)} is
the next clump in the same cell list (0 = end of list).  When a new clump is
accepted (Step~5 above), it is prepended to the head of its cell's list in
$O(1)$.

\paragraph{Why 27 cells are sufficient.}
Two clumps overlap iff their centre separation is $< 2r_c$.  If the
candidate lies in cell $(i,j,k)$, any overlapping clump must satisfy
$|x_\mathrm{cl} - x_c| < 2r_c$.  Since $\ell_\mathrm{RSA} \ge 2r_c$ by
construction (eq.~\ref{eq:ng}), the overlapping clump's cell index $i'$
satisfies $i' \in \{i-1, i, i+1\}$, and likewise for $j'$ and $k'$.
Checking all $3^3 = 27$ cells centred on $(i,j,k)$ is therefore both
sufficient and complete --- no overlapping clump can lie outside this
neighbourhood.

\paragraph{Cell-size constraint.}
The 27-cell rule requires $\ell_\mathrm{RSA} \ge 2r_c$, i.e.\
$N_g \le R/r_c$.  For the reference case ($R = 1$, $r_c = 0.01$):
$R/r_c = 100$, $N_g = \lfloor 66667^{1/3}\rfloor + 1 = 42$,
$\ell_\mathrm{RSA} = 2/42 \approx 0.048 \gg 2r_c = 0.02$. \checkmark

\subsection{Acceptance Rate and Jamming Limit}

As more clumps are placed, their exclusion zones (spheres of radius $2r_c$
around each accepted clump) collectively block an increasing fraction of the
placement domain.  The per-candidate acceptance probability
$p_\mathrm{acc}$ decreases monotonically with $f_\mathrm{vol}$ and
approaches zero at the jamming limit.

For the reference case ($f_\mathrm{cov}=5$, $r_c/R=0.01$, $N=66667$,
$f_\mathrm{vol}=0.067$), the overall acceptance rate is $\approx 74\%$:
roughly 90\,000 candidates are drawn to place 66\,667 clumps.  The algorithm
remains practical for $f_\mathrm{vol} \lesssim 35\%$.

%=============================================================================
\section{CSR Acceleration Grid}
%=============================================================================

\subsection{Motivation}

During radiative transfer, a photon in vacuum must find the nearest clump
along its current flight direction at every propagation step.  Checking all
$N$ clumps per step would cost $O(N)$ per query and become prohibitive for
large populations ($N \sim 10^4$--$10^5$).  The CSR acceleration grid
pre-sorts clumps into spatial cells; the DDA traversal (\S\ref{sec:dda})
then visits only the cells the ray crosses, and within each cell tests only
the few registered clumps.  Together, DDA+CSR achieves $O(1)$ amortised cost
per nearest-clump query, independent of $N$.

\subsection{Data Structure: Compressed Sparse Row}

Instead of a list-of-lists (which requires many small dynamic allocations),
all per-cell clump lists are packed into two flat integer arrays in
\emph{Compressed Sparse Row} (CSR) format --- the same layout used for
sparse matrices:
\begin{sloppypar}
\begin{itemize}
  \item \texttt{cg\_start(0:Ncell)}: \texttt{cg\_start(icell)} is the
        1-based index of the first entry for cell \texttt{icell} in
        \texttt{cg\_list}.  The entries for cell \texttt{icell} occupy
        positions \texttt{cg\_start(icell) \ldots{} cg\_start(icell+1)-1};
        their count is \texttt{cg\_start(icell+1) - cg\_start(icell)}.
  \item \texttt{cg\_list(1:Ntot)}: flat concatenation of all per-cell
        clump-index lists.
\end{itemize}
\end{sloppypar}

Retrieving the clump list for any cell is an $O(1)$ array slice with no
pointer chasing.  To illustrate: suppose four cells contain the clumps
$\{2,5\}$, $\{\}$, $\{1,3\}$, $\{4\}$ respectively:
\begin{align*}
  \texttt{cg\_start} &= [1,\;3,\;3,\;5,\;6], \\
  \texttt{cg\_list}  &= [2,\;5,\;1,\;3,\;4].
\end{align*}
The clumps in cell~2 are \texttt{cg\_list(3:4)} $= \{1,3\}$.
Both arrays reside in MPI-3 shared memory (one copy per compute node).

\subsection{Cell Grid and Domain}
\label{sec:csr_grid}

The CSR grid reuses the same cell count $M = N_g$ as the RSA hash grid
(\S\ref{sec:rsa_hash}), but covers a slightly larger domain
$[-R - r_c,\; R + r_c]^3$ with uniform cell size
\begin{equation}
  \delta = \frac{2(R + r_c)}{M}.
\end{equation}
Expanding the domain by $r_c$ in each direction ensures two properties:
\begin{enumerate}
  \item A clump centred exactly on the outer sphere surface
        ($|\mathbf{c}| = R$) fits entirely inside the grid.
  \item A photon entering from outside the outer sphere starts inside a
        valid grid cell, so the DDA traversal can begin immediately.
\end{enumerate}

\subsection{Clump Registration: AABB Overlap Test}
\label{sec:aabb}

A clump of radius $r_c$ centred at $(c_x, c_y, c_z)$ is registered in every
cell whose axis-aligned bounding box (AABB) overlaps the clump sphere.
Cell $(i,j,k)$ spans $[x_\mathrm{min}+i\,\delta,\; x_\mathrm{min}+(i+1)\,\delta]$
along $x$.  The range of cell indices that overlap $[c_x - r_c,\; c_x + r_c]$ is
\begin{equation}
  i \;\in\; \left[
    \left\lfloor\frac{c_x - r_c - x_\mathrm{min}}{\delta}\right\rfloor,\;
    \left\lfloor\frac{c_x + r_c - x_\mathrm{min}}{\delta}\right\rfloor
  \right],
  \label{eq:aabb_range}
\end{equation}
and likewise for $j$ and $k$.  The clump is registered in every cell in the
Cartesian product of these three index ranges.

For the reference case ($r_c = 0.01$, $\delta \approx 0.048$, so
$r_c/\delta \approx 0.21 < 1$), each clump's index range spans at most 2
values per dimension, so a clump touches at most $2^3 = 8$ cells.  The total
registration count satisfies $N \le N_\mathrm{tot} \le 8N$.

\subsection{Two-Pass Construction}

The CSR arrays are built in two sequential passes over all $N$ clumps,
avoiding any intermediate list storage (the standard technique for CSR
sparse-matrix assembly):

\begin{description}
  \item[Pass 1 (count):] Compute the cell ranges (eq.~\ref{eq:aabb_range})
  for each clump and increment a counter \texttt{cnt(icell)} for every
  touched cell.  Then compute the exclusive prefix sum:
  \[
    \texttt{cg\_start(0)} = 1, \qquad
    \texttt{cg\_start(icell+1)} = \texttt{cg\_start(icell)} + \texttt{cnt(icell)}.
  \]
  Allocate \texttt{cg\_list} of length
  $N_\mathrm{tot} = \texttt{cg\_start(Ncell)} - 1$ as an MPI-3
  shared-memory window.

  \item[Pass 2 (fill):] Repeat the same cell-range loop.  For each touched
  cell, write the clump index at write pointer \texttt{ptr(icell)}, then
  advance the pointer by one.  Initialise
  \texttt{ptr(icell) = cg\_start(icell)} before this pass.
\end{description}

\begin{lstlisting}
!--- Pass 1: count registrations per cell
cnt(:) = 0
do icl = 1, N_clumps
  i0 = max(0,   floor((cl_x(icl) - rc - xmin) / delta))
  i1 = min(M-1, floor((cl_x(icl) + rc - xmin) / delta))
  j0 = ...; j1 = ...                    ! analogous for y
  k0 = ...; k1 = ...                    ! analogous for z
  do k = k0,k1;  do j = j0,j1;  do i = i0,i1
    cnt(cell_idx(i,j,k)) = cnt(cell_idx(i,j,k)) + 1
  enddo;  enddo;  enddo
enddo
! exclusive prefix sum -> cg_start
cg_start(0) = 1
do icell = 0, Ncell-1
  cg_start(icell+1) = cg_start(icell) + cnt(icell)
enddo
Ntot = cg_start(Ncell) - 1
! allocate cg_list(Ntot) as MPI-3 shared-memory window

!--- Pass 2: fill cg_list
ptr(0:Ncell-1) = cg_start(0:Ncell-1)   ! one write pointer per cell
do icl = 1, N_clumps
  ... (same i0,i1,j0,j1,k0,k1 as Pass 1)
  do k = k0,k1;  do j = j0,j1;  do i = i0,i1
    ic = cell_idx(i,j,k)
    cg_list(ptr(ic)) = icl
    ptr(ic) = ptr(ic) + 1
  enddo;  enddo;  enddo
enddo
! invariant: ptr(icell) == cg_start(icell+1) for all cells after Pass 2
\end{lstlisting}

%=============================================================================
\section{Raytrace Through the Clump Medium}
%=============================================================================

\subsection{Overview}

Photon propagation is split into two phases:

\begin{enumerate}
  \item \textbf{Advance through vacuum} -- use the DDA traversal of the CSR
        grid (\S\ref{sec:dda}) to find the next clump hit.
  \item \textbf{Traverse through a clump} -- accumulate optical depth along
        a straight path through the sphere (\S\ref{sec:clump_traverse}).
\end{enumerate}

Frequency shifts due to individual clump bulk velocities are applied at
clump entry and reversed at clump exit.

\subsection{Ray--Sphere Intersection}

Given a ray $\mathbf{p}(t) = \mathbf{o} + t\hat{\mathbf{k}}$ and a sphere
centred at $\mathbf{c}$ with squared radius $r_c^2$, define
$\mathbf{r} = \mathbf{o} - \mathbf{c}$, $b = \mathbf{r}\cdot\hat{\mathbf{k}}$,
$\Delta = b^2 - |\mathbf{r}|^2 + r_c^2$.  If $\Delta < 0$ there is no
intersection.  Otherwise
\begin{equation}
  t_\mathrm{entry} = -b - \sqrt{\Delta},\qquad
  t_\mathrm{exit}  = -b + \sqrt{\Delta}.
\end{equation}
A hit is recorded when $t_\mathrm{exit} > 0$.

\subsection{DDA Search for Next Clump}
\label{sec:dda}

The Amanatides \& Woo (1987) digital differential analyser (DDA) traverses
the CSR cells along the ray in order of increasing path length.  For each
cell, all registered clumps are tested with the ray--sphere formula.  The
traversal stops as soon as the current cell's entry distance exceeds the
best hit distance found so far.

\begin{lstlisting}
best_te  = huge;  best_icl = 0
d        = 0     ! path length to current cell entry

do while (.true.)
   if (d > best_te .or. d > t_max) exit
   ! test all clumps in current cell
   do ip = cg_start(icell), cg_start(icell+1)-1
      icl = cg_list(ip)
      if (icl == skip_icl) cycle
      call ray_sphere_isect(..., te, tx2, hit)
      if (hit .and. tx2>0 .and. te<best_te) then
         best_te=te; best_tx2=tx2; best_icl=icl
      endif
   enddo
   ! advance to next cell (Amanatides & Woo)
   if (tx<=ty .and. tx<=tz) then
      d=tx; ci=ci+si; tx=tx+delx
   elseif (ty<=tz) then
      d=ty; cj=cj+sj; ty=ty+dely
   else
      d=tz; ck=ck+sk; tz=tz+delz
   endif
enddo
\end{lstlisting}

\subsection{Optical Depth Traversal}
\label{sec:clump_traverse}

\texttt{raytrace\_to\_tau\_clump} advances the photon until optical depth
$\tau_\mathrm{in}$ is accumulated or the outer sphere is exited.

\paragraph{Inside a clump ($\texttt{icell\_clump} > 0$):}
The remaining path through the current clump is
$s = t_\mathrm{exit}(\mathbf{p}, \hat{\mathbf{k}}, \mathrm{icl})$.
Opacity at the current frequency:
$\kappa = \kappa_c\,\phi(x, a_V)$.
If $\tau_\mathrm{rem} \le \kappa s$, scatter at $\mathrm{d}s = \tau_\mathrm{rem}/\kappa$.
Otherwise subtract $\kappa s$ from $\tau_\mathrm{rem}$, advance to clump exit,
shift frequency back to the global frame, and continue in vacuum.

\paragraph{In vacuum ($\texttt{icell\_clump} = 0$):}
Call \texttt{find\_next\_clump} (DDA, \S\ref{sec:dda}) to find the nearest
clump.  If none is found along the ray, advance to the sphere boundary and
exit ($\texttt{photon\%inside} = \texttt{.false.}$).

\paragraph{Frequency convention:}
The photon frequency $x$ (in units of $\Delta\nu_D^{cl}$) is maintained in
the local clump rest frame while inside a clump, and in the global frame in
vacuum:
\begin{align}
  \text{at clump entry:}\quad &x \leftarrow x - u_\mathrm{los},\\
  \text{at clump exit:}\quad  &x \leftarrow x + u_\mathrm{los},
\end{align}
where $u_\mathrm{los} = (\mathbf{v}_i \cdot \hat{\mathbf{k}})/v_\mathrm{th}$
is the line-of-sight component of the host clump's bulk velocity in units
of the clump thermal velocity $v_\mathrm{th}=\Delta\nu_D^{cl}\,\lambda_0$.

\paragraph{Internal storage of clump bulk velocities:}
\texttt{cl\_vx/y/z} are stored as $\mathbf{v}/v_\mathrm{th}$ (dimensionless),
matching the Cartesian/AMR convention for \texttt{grid\%vfx} (which is also
pre-normalised to $v/v_\mathrm{th}$ in \texttt{grid\_mod\_car.f90}).  The
clump velocity assignment routines fill the arrays in km/s; \texttt{init\_clumps}
divides by \texttt{cl\_vtherm} after the \texttt{MPI\_BCAST} step so that
all subsequent raytrace and peel-off code can use \texttt{cl\_v*} directly
in $u_\mathrm{los}$ without a per-call division.  The user-facing FITS
output (Section~\ref{sec:clump_fits}) multiplies back by
\texttt{cl\_vtherm} to keep the stored \texttt{VX}/\texttt{VY}/\texttt{VZ}
columns in km/s.

\paragraph{Jout accumulation:}
When the photon exits the outer sphere (\texttt{photon\%inside} becomes
\texttt{.false.}), the photon weight is accumulated into the escape spectrum:
\begin{equation}
  J_\mathrm{out}[i_x] \mathrel{+}= w,\qquad
  i_x = \left\lfloor \frac{x - x_\mathrm{min}}{\delta x}\right\rfloor + 1,
\end{equation}
where $x$ is the photon frequency (global frame, in units of $\Delta\nu_D$)
and $w$ is the photon weight.  This mirrors the behaviour of
\texttt{raytrace\_to\_tau\_car}, where Jout is accumulated upon grid exit.
There are three exit paths (clump-then-sphere exit, already-outside-sphere,
and no-clump-found), all handled identically.

\subsection{Optical Depth to Edge}

\texttt{raytrace\_to\_edge\_clump} computes $\tau$ from the current position
to the sphere boundary without moving the photon (read-only input).  The same
DDA logic is used; frequency shifts at clump entry/exit are tracked locally.
This is called by peel-off routines and by the forced-first-scatter
initialisation in \texttt{run\_simulation\_mod}.

%=============================================================================
\section{Peel-Off Observations}
%=============================================================================

\begin{sloppypar}
Peel-off is handled by the standard \texttt{peelingoff\_rect} routines
(\texttt{peeling\_direct\_outside},
\texttt{peeling\_resonance\_\allowbreak{}nostokes\_outside},
etc.) with \texttt{raytrace\_to\_edge} pointing to
\texttt{raytrace\_to\_edge\_clump}.
\end{sloppypar}

The observer distance defaults to $d = 100\,R$ and the image pixel size is
set so that the full sphere subtends the image:
\begin{equation}
  \delta\theta = \frac{1}{N_x/2}\arcsin\!\left(\frac{R}{d}\right)\;\mathrm{[rad]}
  \approx \frac{R}{d\,(N_x/2)}\;\mathrm{[rad]}.
\end{equation}
For $R=1$, $d=100$, $N_x=129$: $\delta\theta \approx 0\fdg0089$, covering
$\pm 0\fdg57$ (total $1\fdg14$), which exactly spans the sphere.

\paragraph{Clump-aware lab-frame conversion:}
At the moment a peel-off is computed, scattering happens INSIDE a clump,
so \texttt{photon\%xfreq} is in the host clump's rest frame.  The peel-off
output must be in the lab (observer) frame, hence the routines compute
\begin{equation}
  x_\mathrm{ref} = x_\mathrm{photon} + u_1,\qquad
  u_1 = \mathbf{v}_\mathrm{cell}\cdot\hat{\mathbf{k}}_\mathrm{obs}/v_\mathrm{th}.
\end{equation}
For the standard Cartesian path $\mathbf{v}_\mathrm{cell}$ comes from
\texttt{grid\%vfx/y/z}; in clump mode \texttt{grid\%vfx/y/z} are zero
everywhere (the inter-clump medium is vacuum and the bulk velocities live
on the clumps), so the peel-off routines branch on
\texttt{par\%use\_clump\_medium .and. photon\%icell\_clump > 0} and use
the host clump's velocity instead:
\begin{equation}
  u_1 = \texttt{cl\_vx}(\mathrm{icl})\,k_x + \texttt{cl\_vy}(\mathrm{icl})\,k_y + \texttt{cl\_vz}(\mathrm{icl})\,k_z,
\end{equation}
where \texttt{icl}~=~\texttt{photon\%icell\_clump}.  The
\texttt{peeling\_direct\_outside} routine, called only at photon birth
when \texttt{photon\%icell\_clump}~=~0, needs no special case.

Without this branch the peel-off spectra would remain in the clump frame
while \texttt{grid\%Jout} (the angular average) is in the lab frame,
producing a systematically narrower peel-off line profile whenever the
clump bulk velocity field is non-trivial (e.g.\ rotating-halo or
Hubble-flow runs).

\paragraph{FITS WCS for the 3D peel-off cube:}
The peel-off cube is allocated as
\texttt{obs\%scatt(par\%nxfreq, par\%nxim, par\%nyim)}, so the FITS file
has \texttt{NAXIS1}~=~\texttt{nxfreq} (wavelength), \texttt{NAXIS2}~=~\texttt{nxim}
(RA), \texttt{NAXIS3}~=~\texttt{nyim} (DEC).  The header carries a
three-axis WCS:
\begin{center}
\begin{tabular}{l|lll}
\hline
Axis & \texttt{CTYPE} & \texttt{CUNIT} & step \\
\hline
\texttt{NAXIS1} & \texttt{WAVE}    & \texttt{Angstrom} & $\texttt{CD1\_1} = -\,\delta\lambda$ \\
\texttt{NAXIS2} & \texttt{RA-{}-TAN} & \texttt{deg}    & $\texttt{CD2\_2} = \delta\theta_x$ \\
\texttt{NAXIS3} & \texttt{DEC-TAN} & \texttt{deg}      & $\texttt{CD3\_3} = \delta\theta_y$ \\
\hline
\end{tabular}
\end{center}
\texttt{CD1\_1} is negative because $x_\mathrm{freq}$ increases with the
pixel index while wavelength
$\lambda(x) = (1 - v_\mathrm{th}\,x/c)\,\lambda_0$ decreases with~$x$.
\texttt{CRPIX1}~=~$1$, \texttt{CRVAL1}~=~\texttt{grid\%wavelength(1)}
(largest wavelength, at pixel~1).  The scalar header keywords
\texttt{XFREQ1}, \texttt{XFREQ2}, \texttt{DXFREQ}, and \texttt{DWAVE} are
also retained for backward compatibility with existing analysis scripts
that compute the spectral axis from those instead of from the WCS.

%=============================================================================
\section{FITS Output of Clump Information}
\label{sec:clump_fits}
%=============================================================================

When \texttt{par\%save\_clump\_info = .true.}, the subroutine
\texttt{write\_clumps\_fits} (in \texttt{clump\_mod.f90}) writes a binary-table
FITS file \texttt{<base>\_clumps.fits.gz} containing:

\paragraph{BinTable columns:}
\texttt{X}, \texttt{Y}, \texttt{Z} (clump centres, code units);
\texttt{VX}, \texttt{VY}, \texttt{VZ} (bulk velocities, km/s).
The internal arrays \texttt{cl\_vx/y/z} are stored dimensionlessly as
$\mathbf{v}/v_\mathrm{th}$; \texttt{write\_clumps\_fits} multiplies each
column by \texttt{cl\_vtherm} (km/s) before writing, so the user-facing
file always carries velocities in km/s.

\paragraph{Header keywords (in the BinTable HDU):}
\begin{center}
\begin{tabular}{lll}
\hline
Keyword & Value & Description \\
\hline
\texttt{N\_CLUMPS} & $N$ & number of clumps (realised) \\
\texttt{SPHERE\_R} & $R$ & outer sphere radius (code units) \\
\texttt{CL\_RAD}   & $r_c$ & clump radius (code units) \\
\texttt{F\_VOL}    & $f_\mathrm{vol}$ & volume filling factor (realised) \\
\texttt{F\_COV}    & $f_\mathrm{cov}$ & covering factor (realised) \\
\texttt{TAU0}      & $\tau_0$ & line-centre $\tau$ per clump \\
\texttt{SIGMA\_V}  & $\sigma_v$ & bulk velocity dispersion (km/s) \\
\texttt{TEMP\_CL}  & $T_c$ & clump temperature used (K) \\
\texttt{RHOKAP}    & $\kappa_c$ & opacity per code unit \\
\texttt{CL\_DFREQ} & $\Delta\nu_D^{cl}$ & Doppler frequency (Hz) \\
\texttt{VTHERM}    & $v_\mathrm{th}$ & thermal velocity (km/s) \\
\texttt{VOIGT\_A}  & $a_V$ & Voigt damping parameter \\
\texttt{RMAX}      & $R$ & \texttt{par\%rmax} (input) \\
\texttt{IN\_FCOV}  & \texttt{par\%clump\_f\_cov} & input covering factor \\
\texttt{IN\_FVOL}  & \texttt{par\%clump\_f\_vol} & input filling factor \\
\texttt{IN\_NCL}   & \texttt{par\%clump\_N\_clumps} & input $N$ \\
\texttt{IN\_NHI}   & \texttt{par\%clump\_NHI} & per-clump $N_{\rm HI}$ (cm$^{-2}$, clump centre to surface) \\
\texttt{IN\_NH}    & \texttt{par\%clump\_nH} & input $n_H$ (cm$^{-3}$) \\
\texttt{IN\_TEMP}  & \texttt{par\%clump\_temperature} & input $T_c$ (K) \\
\hline
\end{tabular}
\end{center}

Note: column data are written first (which creates the BinTable HDU);
header keywords are then written into that same HDU.  FITS keyword names
must not exceed 8 characters.

%=============================================================================
\section{Input Parameters Summary}
%=============================================================================

\begin{center}
\begin{tabular}{llp{0.55\linewidth}}
\hline
Parameter & Default & Description \\
\hline
\texttt{par\%use\_clump\_medium} & \texttt{.false.} & enable clump mode \\
\texttt{par\%rmax}              & ---  & outer sphere radius (code units) \\
\texttt{par\%clump\_radius}     & ---  & clump radius $r_c$ (code units) \\
\texttt{par\%clump\_f\_cov}     & 0    & covering factor (input) \\
\texttt{par\%clump\_f\_vol}     & 0    & volume filling factor (input) \\
\texttt{par\%clump\_N\_clumps}  & 0    & number of clumps (input, direct) \\
\texttt{par\%clump\_tau0}       & $-1$ & line-centre $\tau$ of one clump (centre to surface) \\
\texttt{par\%clump\_NHI}        & $-1$ & per-clump $N_{\rm HI}$ (cm$^{-2}$, clump centre to surface; no \texttt{distance\_unit} needed) \\
\texttt{par\%clump\_nH}         & $-1$ & clump $n_H$ (cm$^{-3}$; requires \texttt{distance\_unit}) \\
\texttt{par\%clump\_temperature}& $-1$ & clump $T$ (K); $<0$ $\Rightarrow$ \texttt{par\%temperature} \\
\texttt{par\%clump\_sigma\_v}   & 0    & random velocity $\sigma$ per component (km/s) \\
\texttt{par\%velocity\_type}    & \texttt{''} & systematic flow type (see \S2.4) \\
\texttt{par\%Vexp}              & 0    & expansion speed for \texttt{hubble}/\texttt{constant\_radial} (km/s) \\
\texttt{par\%Vx, Vy, Vz}        & 0    & bulk velocity for \texttt{parallel\_velocity} (km/s) \\
\texttt{par\%Vpeak}             & 0    & peak speed for \texttt{ssh} (km/s) \\
\texttt{par\%rpeak}             & 0    & peak radius for \texttt{ssh} (code units) \\
\texttt{par\%DeltaV}            & 0    & outer velocity increment for \texttt{ssh} (km/s) \\
\texttt{par\%Vrot}              & 0    & rotation speed for rotating models (km/s) \\
\texttt{par\%rinner}            & 0    & solid-body radius for \texttt{rotating\_galaxy\_halo} \\
\texttt{par\%save\_clump\_info} & \texttt{.false.} & write clump FITS file \\
\hline
\end{tabular}
\end{center}

\paragraph{Example input:}
\begin{lstlisting}
&parameters
 par%use_clump_medium = .true.
 par%rmax             = 1.0
 par%clump_radius     = 0.01
 par%clump_f_cov      = 5.0
 par%clump_tau0       = 1000.0
 par%temperature      = 1.0e4
 par%clump_sigma_v    = 0.0
 par%no_photons       = 1e5
 par%spectral_type    = 'voigt'
 par%nxfreq           = 301
 par%velocity_min     = -100.0
 par%velocity_max     =  100.0
 par%nx = 11, par%ny = 11, par%nz = 11
 par%nxim = 129, par%nyim = 129
 par%save_clump_info  = .true.
 par%out_file         = 'clump_test.fits.gz'
/
\end{lstlisting}

%=============================================================================
\section{Memory Layout}
%=============================================================================

\begin{sloppypar}
Clump positions, velocities, and CSR arrays are allocated as MPI-3
shared-memory windows via \texttt{create\_shared\_mem} from
\texttt{memory\_mod}, giving one copy per compute node rather than per MPI
rank.  On systems with multiple ranks per node this significantly reduces
memory usage for large clump populations.
\end{sloppypar}

\begin{sloppypar}
Cleanup is performed by \texttt{destroy\_clumps} which calls
\texttt{destroy\_shared\_mem\_all} (collective \texttt{MPI\_WIN\_FREE} across
all ranks) followed by explicit \texttt{nullify} of all pointers.  The
individual \texttt{destroy\_mem} is not used because
\texttt{MPI\_WIN\_SHARED\_QUERY} returns the rank-0 base address, which
differs from the local pointer on non-root ranks.
\end{sloppypar}

%=============================================================================
\section{Validation and Examples}
%=============================================================================

\begin{sloppypar}
The directory \texttt{examples/clump\_sphere/} contains a suite of six
reference runs (plus the legacy \texttt{clump\_test.in}) covering the
main parameter space.  All named cases use $R=1$, $r_c=0.01$, $10^6$
photons, and \texttt{par\%use\_clump\_medium = .true.}:
\end{sloppypar}

\begin{center}
\begin{tabular}{lll}
\hline
Input file & $\tau_0$ & Notes \\
\hline
\texttt{clump\_tau3\_fcov5.in}        & $10^3$ & $f_\mathrm{cov}=5$, static (reference) \\
\texttt{clump\_tau4\_fcov5.in}        & $10^4$ & $f_\mathrm{cov}=5$, static \\
\texttt{clump\_tau3\_fcov20.in}       & $10^3$ & $f_\mathrm{cov}=20$, dense covering \\
\texttt{clump\_tau3\_fcov5\_hubble.in}& $10^3$ & $f_\mathrm{cov}=5$, Hubble $V_\mathrm{exp}=200$ km/s \\
\texttt{clump\_tau3\_fcov5\_sigv.in}  & $10^3$ & $f_\mathrm{cov}=5$, $\sigma_v=50$ km/s \\
\texttt{clump\_tau3\_fcov5\_peel.in}  & $10^3$ & $f_\mathrm{cov}=5$, peel-off $129\times129$ \\
\hline
\end{tabular}
\end{center}

Run with \texttt{bash run.sh} (uses 4 MPI ranks by default) or individually:
\begin{verbatim}
mpirun -np 4 ../../LaRT.x clump_tau3_fcov5.in
\end{verbatim}
Python post-processing:
\begin{verbatim}
python plot_clump.py          # all runs
python plot_clump.py clump_tau3_fcov5   # single run
\end{verbatim}

For the reference case ($f_\mathrm{cov}=5$, $\tau_0=10^3$):
\begin{itemize}
  \item Realised $f_\mathrm{cov} \approx 5.00$, $f_\mathrm{vol} \approx 0.067$;
  \item RSA acceptance rate $\sim 74\%$, $N=66667$ clumps placed;
  \item Mean scattering count $\approx 6400$ per photon;
  \item Wall time $\approx 3.75$ min (10 ranks, $10^5$ photons).
\end{itemize}

The escaped spectrum $J_\mathrm{out}(x)$ shows the characteristic double-peak
expected for a clumpy Lyman-$\alpha$ source.  The peel-off output
(\texttt{\_obs.fits.gz}) contains \texttt{Scattered} (HDU~0) and
\texttt{Direct} (HDU~1) spectral image cubes of shape $(301, 129, 129)$.

\end{document}
